---
title: "pH_Relative_CH4Conc_Difference"
author: "Ellery Vaughan"
date: "2025-10-06"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3$
    toc_float: no
    df_print: kable 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# pH Experiment CH4 & CO2 Concentration Analysis 

**Script Notes**

PREVIOUS ANALYSIS ATTEMPTS:
1) I also repeated this analysis for the concentration values but normalized by incubation duration time (days). It did not change the significance of the results for timepoint 2, but the results were a little different for timepoint 1, but made less sense. So i am just going with raw concentration values because I think they're more directly interpretable. Time normalized concentration analysis are now archived in old section of code (unless i really decide i don't need later and decide to delete.)
2) Also moved code to OLD for combining the T1 & T2 plots into one figure because we decided we didn't want to present in 4 panel way because didn't want to encourage the comparisons between T1 and T2 concentrations within soil type. 
3) I made figures / did the analysis combining the pH treatments with the wet control from the H2O experiment, but the concentrations were so much lower for the wet control jars. The comparison didn't really make sense. So moved this code to the OLD section. 
4) Lastly, I'm going to do the concentration analysis including the dry control to double check what that looks like before committing to just dropping the dry control and moving forward with the 5, 7 & 8 treatments 


## Set up 
```{r loading libraries}

library(tidyverse)
library(sjPlot)
library(ggtext)
library(ggh4x)
library(readxl)
library(ggforce)
library(dplyr)
library(purrr)
library(emmeans)
library(multcomp)        # not strictly required, but helpful
library(multcompView)    # supplies helper functions used by cld()
library(wesanderson)

```

```{r reading in and cleaning data}

# concentration data 
raw <- readxl::read_xlsx("data/EV_pH_13CH4Inc_CH4_CO2_10.21.25.xlsx",
                          sheet = "CH4_CO2_conc_results")

data <- raw[1:46,1:18]

data <- data %>%
  mutate(
    # Convert to factors and relevel
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8")),
    rep          = factor(rep, levels = c("1", "2", "3", "4", "5")),
    jar_type     = as.factor(jar_type),
    
    # Convert multiple columns to numeric using across()
    across(
      c(CH4_ppm_FINAL, CO2_ppm_INITIAL, CO2_ppm_FINAL,
        T0_to_T1_days, T1_to_T2_days, CH4_CV_F),
      as.numeric),
    
    # adding columns for time normalized concentrations 
    CH4_ppm_I_timenorm = CH4_ppm_INITIAL/T0_to_T1_days,
    CH4_ppm_F_timenorm = CH4_ppm_FINAL/T1_to_T2_days,
    CO2_ppm_I_timenorm = CO2_ppm_INITIAL/T0_to_T1_days,
    CO2_ppm_F_timenorm = CO2_ppm_FINAL/T1_to_T2_days    
    )

# write_csv(data,"all_data.csv")

```

# CH4 concentration analyses (without dry control)

```{r CH4 - summarizing data for CH4 concentration figures (initial & final) }

# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jars <- data %>%
  filter(
    jar_type == "Exp",
    pH_treatment != "Dry control"
  )

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CH4_ppm_INITIAL, CH4_ppm_FINAL, CH4_ppm_I_timenorm, CH4_ppm_F_timenorm),  # Select the numeric columns to summarize: initial and final CH4 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CH4_ppm_INITIAL = CH4_ppm_INITIAL_sd / sqrt(CH4_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CH4
    se_CH4_ppm_FINAL   = CH4_ppm_FINAL_sd / sqrt(CH4_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CH4
    se_CH4_ppm_I_timenorm   = CH4_ppm_I_timenorm_sd / sqrt(CH4_ppm_I_timenorm_n),
    se_CH4_ppm_F_timenorm   = CH4_ppm_F_timenorm_sd / sqrt(CH4_ppm_F_timenorm_n),
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  
  )

# write_csv(summ_exp,"summarized_data.csv")

# making table to copy values into summary table for mauscript SI
CH4table_summ_exp <- data %>% 
  filter(jar_type == "Exp") %>% 
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CH4_ppm_INITIAL, CH4_ppm_FINAL),  # Select the numeric columns to summarize: initial and final CH4 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CH4_ppm_INITIAL = CH4_ppm_INITIAL_sd / sqrt(CH4_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CH4
    se_CH4_ppm_FINAL   = CH4_ppm_FINAL_sd / sqrt(CH4_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CH4
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("Dry control","5", "7", "8"))  
  ) %>% 
  dplyr::select(soil_type,
         pH_treatment,
         CH4_ppm_INITIAL_mean,
         se_CH4_ppm_INITIAL,
         CH4_ppm_FINAL_mean,
         se_CH4_ppm_FINAL)

write_csv(CH4table_summ_exp,"manuscript_materials/pH_CH4ppm_summary.csv")



```

## Timepoint 1 (initial)

```{r TIMEPOINT 1 CH4 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting}

# 1) linear models
# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type

CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_INITIAL_mean, se_CH4_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL + 0.05*max(CH4_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# 3) 
# Create p-value annotations from model summaries
CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 4) Plot CH4 conc means and lm stats results

# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_INITIAL_mean - se_CH4_ppm_INITIAL,
                    ymax = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_I, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/pH/concentrations/pH_CH4_conc_INITIAL_nocontrol.png", width = 8.5, height = 5)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 CH4 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting }

# 1) linear models
# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  emmeans_res <- CH4_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8") )  # Relevel pH_treatment
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type
CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_FINAL_mean, se_CH4_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL + 0.075*max(CH4_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 3) Create p-value annotations from model summaries
CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 4) plot CH4 conc means and lm stats results
# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_FINAL_mean - se_CH4_ppm_FINAL,
                    ymax = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_F, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/pH/concentrations/pH_CH4_conc_FINAL_nocontrol.png", width = 8.5, height = 5)

```

# CO2 concentration analyses (without dry control)

```{r CO2 - summarizing data for CO2 concentration figures (initial & final) }

# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jars <- data %>%
  filter(
    jar_type == "Exp",
    pH_treatment != "Dry control"
  )

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CO2_ppm_INITIAL, CO2_ppm_FINAL, CO2_ppm_I_timenorm, CO2_ppm_F_timenorm),  # Select the numeric columns to summarize: initial and final CO2 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CO2_ppm_INITIAL_mean, CO2_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CO2_ppm_INITIAL = CO2_ppm_INITIAL_sd / sqrt(CO2_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CO2
    se_CO2_ppm_FINAL   = CO2_ppm_FINAL_sd / sqrt(CO2_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CO2
    se_CO2_ppm_I_timenorm   = CO2_ppm_I_timenorm_sd / sqrt(CO2_ppm_I_timenorm_n),
    se_CO2_ppm_F_timenorm   = CO2_ppm_F_timenorm_sd / sqrt(CO2_ppm_F_timenorm_n),
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  
  )

# write_csv(summ_exp,"summarized_data.csv")

```

## Timepoint 1 (initial)

```{r TIMEPOINT 1 CO2 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting}

# 1) linear models
# Fit models for each soil_type and store results
CO2_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CO2_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CO2 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CO2_models

# Unamended soils model results 
summary(CO2_models_I$Unamended$model) # model summary 
anova(CO2_models_I$Unamended$model) # ANOVA table 
CO2_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CO2_models_I$`Ground Rock`$model) # model summary 
anova(CO2_models_I$`Ground Rock`$model) # ANOVA table 
CO2_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CO2_letters_I <- map_df(names(CO2_models_I), function(st) {
  emmeans_res <- CO2_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type}
CO2_letters_I <- CO2_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CO2_ppm_INITIAL_mean, se_CO2_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CO2_ppm_INITIAL_mean + se_CO2_ppm_INITIAL + 0.05*max(CO2_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# 3) Create p-value annotations from model summaries
CO2_pvals_I <- map_df(names(CO2_models_I), function(st) {
  model <- CO2_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

# 4) plot CH4 conc means and lm stats results
# Plot
CO2_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CO2_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CO2_ppm_INITIAL_mean - se_CO2_ppm_INITIAL,
                    ymax = CO2_ppm_INITIAL_mean + se_CO2_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CO2_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CO2_pvals_I, 
    aes(x = 3, 
        # y = 175500,
        y = max(summ_exp$CO2_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CO[2]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CO[2]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) 
# +
 # ylim(0,176000)

CO2_conc_means_I

ggsave(plot = CO2_conc_means_I, filename = "figures/pH/pH_CO2_conc_INITIAL_mean_yvary.png", width = 8.5, height = 5)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 CO2 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting}

# 1) linear models
# Fit models for each soil_type and store results
CO2_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CO2_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CO2 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CO2_models

# Unamended soils model results 
summary(CO2_models_F$Unamended$model) # model summary 
anova(CO2_models_F$Unamended$model) # ANOVA table 
CO2_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CO2_models_F$`Ground Rock`$model) # model summary 
anova(CO2_models_F$`Ground Rock`$model) # ANOVA table 
CO2_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CO2_letters_F <- map_df(names(CO2_models_F), function(st) {
  emmeans_res <- CO2_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type
CO2_letters_F <- CO2_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CO2_ppm_FINAL_mean, se_CO2_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CO2_ppm_FINAL_mean + se_CO2_ppm_FINAL + 0.05*max(CO2_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

# 3) Create p-value annotations from model summaries
CO2_pvals_F <- map_df(names(CO2_models_F), function(st) {
  model <- CO2_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

# 4) plot CO2 conc means and lm stats results
# Plot
CO2_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CO2_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CO2_ppm_FINAL_mean - se_CO2_ppm_FINAL,
                    ymax = CO2_ppm_FINAL_mean + se_CO2_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CO2_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CO2_pvals_F, 
    aes(x = 3, 
        y = 175500,
        # y = max(summ_exp$CO2_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CO[2]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CO[2]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,176000)

CO2_conc_means_F

ggsave(plot = CO2_conc_means_F, filename = "figures/pH/pH_CO2_conc_FINAL_mean.png", width = 8.5, height = 5)

```


# CH4 concentration analyses (WITH DRY control)

```{r CH4 - summarizing data for CH4 concentration figures (initial & final) }

# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jars <- data %>%
  filter(
    jar_type == "Exp"
  )

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CH4_ppm_INITIAL, CH4_ppm_FINAL, CH4_ppm_I_timenorm, CH4_ppm_F_timenorm),  # Select the numeric columns to summarize: initial and final CH4 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CH4_ppm_INITIAL = CH4_ppm_INITIAL_sd / sqrt(CH4_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CH4
    se_CH4_ppm_FINAL   = CH4_ppm_FINAL_sd / sqrt(CH4_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CH4
    se_CH4_ppm_I_timenorm   = CH4_ppm_I_timenorm_sd / sqrt(CH4_ppm_I_timenorm_n),
    se_CH4_ppm_F_timenorm   = CH4_ppm_F_timenorm_sd / sqrt(CH4_ppm_F_timenorm_n),
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8"))  
  )

```

## Timepoint 1 (initial)

```{r TIMEPOINT 1 CH4 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting}

# 1) linear models
# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8"))  # Relevel pH_treatment
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type

CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_INITIAL_mean, se_CH4_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL + 0.05*max(CH4_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8"))  # Relevel pH_treatment
  )

# 3) 
# Create p-value annotations from model summaries
CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 4) Plot CH4 conc means and lm stats results

# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_INITIAL_mean - se_CH4_ppm_INITIAL,
                    ymax = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_I, 
    aes(x = 1, 
        y = 7.3,
        # y = max(summ_exp$CH4_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  # scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  scale_fill_manual(values = c("darkgrey", "#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values=wes_palette("Darjeeling2", 4) ) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.5)

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/pH/concentrations/pH_CH4_conc_INITIAL_drycontrol.png", width = 8.5, height = 5)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 CH4 - 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting }

# 1) linear models
# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2) Post-hoc pairwise comparisons and generation of compact letter displays
#    This code computes Tukey-style group letters (a, b, ab, etc.) for each soil type
#    after running the CH4 models stored in CH4_models_F.

CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  
  # ---- Step 1: Compute estimated marginal means (EMMs) ----
  # For each soil type (st), extract its model and calculate means by pH_treatment.
  emm <- emmeans(CH4_models_F[[st]]$model, ~ pH_treatment)
  
  # ---- Step 2: Generate compact letter display (CLD) ----
  # The CLD groups treatments that are not significantly different (e.g., "a", "ab", "b").
  cld_df <- multcomp::cld(emm, Letters = letters)
  
  # ---- Step 3: Convert to a plain data frame ----
  # cld() returns an S4 object; converting it allows use of dplyr functions.
  cld_df <- as.data.frame(cld_df)
  
  # ---- Step 4: Select and reorder relevant columns ----
  # Keep only the treatment name and grouping column, 
  # and make sure pH_treatment follows the desired left-to-right order.
  cld_df <- cld_df %>%
    dplyr::select(pH_treatment, .group) %>%
    dplyr::mutate(
      pH_treatment = factor(pH_treatment,
                            levels = c("Dry control", "5", "7", "8"))
    ) %>%
    arrange(pH_treatment)
  
  # ---- Step 5: Enforce left-to-right alphabetical letter assignment ----
  # By default, cld() may assign letters in arbitrary order (e.g., "b" left of "a").
  # This block forces consistent alphabetical ordering across panels.
  unique_groups <- unique(cld_df$.group)              # Identify unique letter groups (e.g., "a", "ab", "b")
  forced_letters <- letters[seq_along(unique_groups)] # Create ordered letters (a, b, c, ...)
  map_groups <- setNames(forced_letters, unique_groups) # Map original groups to ordered letters
  cld_df$letters <- map_groups[cld_df$.group]         # Replace with ordered version
  
  # ---- Step 6: Add soil type identifier ----
  # Store the soil type this result came from, so all results can be combined.
  cld_df$soil_type <- st
  
  # ---- Step 7: Keep only the relevant columns ----
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]
  
  # Return this table for the current soil type
  cld_df
}) %>%
  # ---- Step 8: Final factor level adjustments ----
  # Ensure consistent ordering for soil_type and pH_treatment across all combined data.
  dplyr::mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8"))
  )

# dynamically assign posthoc pairwise letter positions based on max value per soil type
CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_FINAL_mean, se_CH4_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL + 0.075*max(CH4_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

# 3) Create p-value annotations from model summaries
CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 1))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

# 4) plot CH4 conc means and lm stats results
# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_FINAL_mean - se_CH4_ppm_FINAL,
                    ymax = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_F, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  # scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  scale_fill_manual(values = c("darkgrey", "#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values=wes_palette("Darjeeling2", 4) ) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/pH/concentrations/pH_CH4_conc_FINAL_drycontrol.png", width = 8.5, height = 5)

```


# OLD CODE 

# CH4 concentration analyses (WITH WET control)

```{r CH4 - summarizing data for concentration figures (initial & final) }

# pH Experiment 
# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jarspH <- data %>%
  filter(
    jar_type == "Exp",
    pH_treatment != "Dry control"
  ) %>% 
  dplyr::select(jar_num:sample_code,CH4_ppm_INITIAL:CO2_ppm_FINAL)

# H2O Experiment 
# concentration data 
raw <- readxl::read_xlsx("data/EV_H2O_13CH4Inc_CH4_CO2_10.21.25.xlsx",
                          sheet = "CH4_CO2_conc_results")
data <- raw[1:35,1:18]

data <- data %>%
  mutate(
    # Convert to factors and relevel
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),
    H2O_treatment = factor(H2O_treatment, levels = c("no","yes") ),
    rep          = factor(rep, levels = c("1", "2", "3", "4", "5")),
    jar_type     = as.factor(jar_type),
    
    # Convert multiple columns to numeric using across()
    across(
      c(CH4_ppm_FINAL, CO2_ppm_FINAL,
        T0_to_T2_days, T1_to_T2_days, CH4_CV_F),
      as.numeric),
    
    # adding columns for time normalized concentrations 
    )

exp_jarsH2O <- data %>%
  filter(
    jar_type == "Exp",
    H2O_treatment == "yes" 
  ) %>% 
  mutate(H2O_treatment = recode(H2O_treatment, "yes" = "Wet control") ) %>% 
  rename(pH_treatment = H2O_treatment) %>% 
  dplyr::select(jar_num:sample_code,CH4_ppm_INITIAL:CO2_ppm_FINAL)

# combining data from pH and H2O experiments 
exp_jars <- rbind(exp_jarspH,exp_jarsH2O)

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CH4_ppm_INITIAL, CH4_ppm_FINAL),  # Select the numeric columns to summarize: initial and final CH4 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CH4_ppm_INITIAL = CH4_ppm_INITIAL_sd / sqrt(CH4_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CH4
    se_CH4_ppm_FINAL   = CH4_ppm_FINAL_sd / sqrt(CH4_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CH4
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("Wet control","5", "7", "8"))  
  )

# write_csv(summ_exp,"summarized_data.csv")


```

## Timepoint 1 (initial)

```{r 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting }

# 1) making linear models 
# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2a) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Wet control","5", "7", "8"))  # Relevel pH_treatment
  )

# 2b) dynamically assign posthoc pairwise letter positions based on max value per soil type
CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_INITIAL_mean, se_CH4_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL + 0.05*max(CH4_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Wet control", "5", "7", "8"))  # Relevel pH_treatment
  )

#  3) Create p-value annotations from model summaries
CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 4) plot CH4 conc means and lm stats results
# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_INITIAL_mean - se_CH4_ppm_INITIAL,
                    ymax = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_I, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values=wes_palette("Darjeeling2", 4) ) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/pH/concentrations/pH_CH4_conc_INITIAL_wetcontrol.png", width = 8.5, height = 5)

```

## Timepoint 2 (final)

```{r 1) linear modeling, 2) posthoc pairwise comparison, 3) assigning p-values, and 4) plotting }

# 1) making linear models 
# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              

# 2a) post-hoc pairwise comparisons and dynamically generate letters
# Generate compact letter display for each soil type
CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  emmeans_res <- CH4_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Wet control","5", "7", "8"))  # Relevel pH_treatment
  )

# 2b) dynamically assign posthoc pairwise letter positions based on max value per soil type
CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_FINAL_mean, se_CH4_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL + 0.05*max(CH4_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("Wet control", "5", "7", "8"))  # Relevel pH_treatment
  )

#  3) Create p-value annotations from model summaries
CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )


# 4) plot CH4 conc means and lm stats results
# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_FINAL_mean - se_CH4_ppm_FINAL,
                    ymax = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_F, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#046C9A", "#D69C4E","#ABDDDE")) +
  # scale_fill_manual(values=wes_palette("Darjeeling2", 4) ) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/pH/concentrations/pH_CH4_conc_FINAL_wetcontrol.png", width = 8.5, height = 5)

```

## Combining timepoints 1 & 2 into one figrue 

```{r combined T1 & T2 CH4 conc figures}

# 1 Reshape summary data for plotting
summ_all <- summ_exp %>%
  dplyr::select(
    soil_type, pH_treatment,
    CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_mean,
    se_CH4_ppm_INITIAL, se_CH4_ppm_FINAL
  ) %>%
  pivot_longer(
    cols = -c(soil_type, pH_treatment),
    names_to = c("metric", "timepoint"),
    names_pattern = "(CH4_ppm|se_CH4_ppm)_(INITIAL|FINAL)"
  ) %>%
  mutate(
    type = if_else(str_detect(metric, "^se"), "se", "mean"),
    timepoint = recode(timepoint,
                       "INITIAL" = "Timepoint 1 (Initial)",
                       "FINAL"   = "Timepoint 2 (Final)")
  ) %>%
  dplyr::select(-metric) %>%
  pivot_wider(names_from = type, values_from = value)

# 2 Generate compact letters
generate_letters <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    emmeans_res <- CH4_models[[st]]$emmeans$emmeans
    cld_df <- multcomp::cld(emmeans_res, Letters = letters) %>%
      as.data.frame() %>%
      mutate(
        soil_type = st,
        timepoint = timepoint_label,
        letters = .group
      ) %>%
      dplyr::select(soil_type, pH_treatment, timepoint, letters)
  }) %>%
    mutate(
      soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),
      pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))
    )
}

CH4_letters_all <- bind_rows(
  generate_letters(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_letters(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "pH_treatment", "timepoint")) %>%
  mutate(letter_y = mean + se + 0.1 * max(mean))

# 3 Generate p-values
generate_pvals <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    model <- CH4_models[[st]]$model
    pval <- anova(model)$`Pr(>F)`[1]
    data.frame(
      soil_type = st,
      timepoint = timepoint_label,
      label = paste0("p = ", signif(pval, 3))
    )
  }) %>%
    mutate(soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")))
}

CH4_pvals_all <- bind_rows(
  generate_pvals(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_pvals(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "timepoint")) %>%
  group_by(soil_type, timepoint) %>%
  summarise(
    y = 7.7,
    # y = max(mean) + 0.15 * max(mean),
    label = unique(label),
    .groups = "drop"
  )

# 4 Plot combined figure
CH4_conc_combined_plot <- ggplot(summ_all, aes(x = pH_treatment, y = mean, fill = pH_treatment)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, size = 0.3) +
  geom_text(
    data = CH4_letters_all,
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_all,
    aes(x = Inf, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 0,
    size = 4.2,
    # fontface = "italic"
  ) +
  facet_grid(soil_type ~ timepoint) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62")) +
  labs(
    x = "pH Treatment",
    y = expression(paste(CH[4], " Concentration (ppm)")),
    title = expression(bold("Methane Concentrations Across Soil Types and Timepoints"))
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  ylim(0,8)

CH4_conc_combined_plot 

# Display and save
CH4_conc_combined_plot
ggsave(plot = CH4_conc_combined_plot,
       filename = "figures/10.13.25/CH4_conc_combined_timepoints.png",
       width = 10, height = 7)



```

# Time normalized CH4 concentration analyses (without dry control)

## Timepoint 1 (initial)

```{r TIMEPOINT 1 - linear models}

# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_I_timenorm ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 1 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_I_timenorm_mean, se_CH4_ppm_I_timenorm),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_I_timenorm_mean + se_CH4_ppm_I_timenorm + 0.05*max(CH4_ppm_I_timenorm_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - Create p-value annotations from model summaries}

CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 3))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 1 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_I_timenorm_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_I_timenorm_mean - se_CH4_ppm_I_timenorm,
                    ymax = CH4_ppm_I_timenorm_mean + se_CH4_ppm_I_timenorm),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(data = CH4_pvals_I, aes(x = 3, y = max(summ_exp$CH4_ppm_I_timenorm_mean)*1.1, label = label),
             inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Conc~Normalized~by~Incubation~Duration~(Days)~T1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/10.13.25/CH4_conc_normalized_INITIAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 - linear models}

# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_F_timenorm ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 2 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  emmeans_res <- CH4_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8") )  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 2 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_F_timenorm_mean, se_CH4_ppm_F_timenorm),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_F_timenorm_mean + se_CH4_ppm_F_timenorm + 0.05*max(CH4_ppm_F_timenorm_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - Create p-value annotations from model summaries}

CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 3))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_F_timenorm_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_F_timenorm_mean - se_CH4_ppm_F_timenorm,
                    ymax = CH4_ppm_F_timenorm_mean + se_CH4_ppm_F_timenorm),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(data = CH4_pvals_F, aes(x = 3, y = max(summ_exp$CH4_ppm_F_timenorm_mean)*1.1, label = label),
             inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Conc~Normalized~by~Incubation~Duration~(Days)~T2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/10.13.25/CH4_conc_normalized_FINAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

## Combining timepoints 1 & 2 into one figrue 

```{r combined T1 & T2 CH4 conc figures}

# 1 Reshape summary data for plotting
summ_all <- summ_exp %>%
  dplyr::select(
    soil_type, pH_treatment,
    CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_mean,
    se_CH4_ppm_INITIAL, se_CH4_ppm_FINAL
  ) %>%
  pivot_longer(
    cols = -c(soil_type, pH_treatment),
    names_to = c("metric", "timepoint"),
    names_pattern = "(CH4_ppm|se_CH4_ppm)_(INITIAL|FINAL)"
  ) %>%
  mutate(
    type = if_else(str_detect(metric, "^se"), "se", "mean"),
    timepoint = recode(timepoint,
                       "INITIAL" = "Timepoint 1 (Initial)",
                       "FINAL"   = "Timepoint 2 (Final)")
  ) %>%
  dplyr::select(-metric) %>%
  pivot_wider(names_from = type, values_from = value)

# 2 Generate compact letters
generate_letters <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    emmeans_res <- CH4_models[[st]]$emmeans$emmeans
    cld_df <- multcomp::cld(emmeans_res, Letters = letters) %>%
      as.data.frame() %>%
      mutate(
        soil_type = st,
        timepoint = timepoint_label,
        letters = .group
      ) %>%
      dplyr::select(soil_type, pH_treatment, timepoint, letters)
  }) %>%
    mutate(
      soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),
      pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))
    )
}

CH4_letters_all <- bind_rows(
  generate_letters(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_letters(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "pH_treatment", "timepoint")) %>%
  mutate(letter_y = mean + se + 0.1 * max(mean))

# 3 Generate p-values
generate_pvals <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    model <- CH4_models[[st]]$model
    pval <- anova(model)$`Pr(>F)`[1]
    data.frame(
      soil_type = st,
      timepoint = timepoint_label,
      label = paste0("p = ", signif(pval, 3))
    )
  }) %>%
    mutate(soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")))
}

CH4_pvals_all <- bind_rows(
  generate_pvals(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_pvals(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "timepoint")) %>%
  group_by(soil_type, timepoint) %>%
  summarise(
    y = 7.7,
    # y = max(mean) + 0.15 * max(mean),
    label = unique(label),
    .groups = "drop"
  )

# 4 Plot combined figure
CH4_conc_combined_plot <- ggplot(summ_all, aes(x = pH_treatment, y = mean, fill = pH_treatment)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, size = 0.3) +
  geom_text(
    data = CH4_letters_all,
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_all,
    aes(x = Inf, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 0,
    size = 4.2,
    # fontface = "italic"
  ) +
  facet_grid(soil_type ~ timepoint) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62")) +
  labs(
    x = "pH Treatment",
    y = expression(paste(CH[4], " Concentration (ppm)")),
    title = expression(bold("Methane Concentrations Across Soil Types and Timepoints"))
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  ylim(0,8)

CH4_conc_combined_plot 

# Display and save
CH4_conc_combined_plot
ggsave(plot = CH4_conc_combined_plot,
       filename = "figures/10.13.25/CH4_conc_combined_timepoints.png",
       width = 10, height = 7)

```

