---
title: "pH_Relative_CH4Conc_Difference"
author: "Ellery Vaughan"
date: "2025-10-06"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3$
    toc_float: no
    df_print: kable 
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# pH Experiment CH4 & CO2 Concentration Analysis 

**Notes on script iterations & analysis process:**
I also repeated this analysis for the concentration values but normalized by incubation duration time (days). It did not change the significance of the results for timepoint 2, but the results were a little different for timepoint 1, but made less sense. So i am just going with raw concentration values because I think they're more directly interpretable. 
Time normalized concentration analysis are now archived in old section of code (unless i really decide i don't need later and decide to delete.)
Also moved code to OLD for combining the T1 & T2 plots into one figure because we decided we didn't want to present in 4 panel way because didn't want to encourage the comparisons between T1 and T2 concentrations within soil type. 

## Set up 
```{r loading libraries}

library(tidyverse)
library(sjPlot)
library(ggtext)
library(ggh4x)
library(readxl)
library(ggforce)
library(dplyr)
library(purrr)
library(emmeans)
library(multcomp)        # not strictly required, but helpful
library(multcompView)    # supplies helper functions used by cld()

```

```{r reading in and cleaning data}

# concentration data 
raw <- readxl::read_xlsx("data/EV_pH_13CH4Inc_CH4_CO2_10.07.25.xlsx",
                          sheet = "CH4_CO2_conc_results")

data <- raw[1:46,1:18]

data <- data %>%
  mutate(
    # Convert to factors and relevel
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),
    pH_treatment = factor(pH_treatment, levels = c("Dry control", "5", "7", "8")),
    rep          = factor(rep, levels = c("1", "2", "3", "4", "5")),
    jar_type     = as.factor(jar_type),
    
    # Convert multiple columns to numeric using across()
    across(
      c(CH4_ppm_FINAL, CO2_ppm_INITIAL, CO2_ppm_FINAL,
        T0_to_T1_days, T1_to_T2_days, CH4_CV_F),
      as.numeric),
    
    # adding columns for time normalized concentrations 
    CH4_ppm_I_timenorm = CH4_ppm_INITIAL/T0_to_T1_days,
    CH4_ppm_F_timenorm = CH4_ppm_FINAL/T1_to_T2_days,
    CO2_ppm_I_timenorm = CO2_ppm_INITIAL/T0_to_T1_days,
    CO2_ppm_F_timenorm = CO2_ppm_FINAL/T1_to_T2_days    
    )

# write_csv(data,"all_data.csv")

```

# CH4 concentration analyses (without dry control)

```{r CH4 - summarizing data for concentration figures (initial & final) }

# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jars <- data %>%
  filter(
    jar_type == "Exp",
    pH_treatment != "Dry control"
  )

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CH4_ppm_INITIAL, CH4_ppm_FINAL, CH4_ppm_I_timenorm, CH4_ppm_F_timenorm),  # Select the numeric columns to summarize: initial and final CH4 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CH4_ppm_INITIAL = CH4_ppm_INITIAL_sd / sqrt(CH4_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CH4
    se_CH4_ppm_FINAL   = CH4_ppm_FINAL_sd / sqrt(CH4_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CH4
    se_CH4_ppm_I_timenorm   = CH4_ppm_I_timenorm_sd / sqrt(CH4_ppm_I_timenorm_n),
    se_CH4_ppm_F_timenorm   = CH4_ppm_F_timenorm_sd / sqrt(CH4_ppm_F_timenorm_n),
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  
  )

# write_csv(summ_exp,"summarized_data.csv")


```

## Timepoint 1 (initial)

```{r TIMEPOINT 1 - linear models}

# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 1 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r Quick summary:}
# 🧠 Quick summary:
# 
# map_df() → iterates over all soil types and combines results into one data frame automatically
# 
# emmeans_res → pulls the estimated marginal means table for the current soil type
# 
# cld() → computes compact letters showing which treatment means are statistically different
# 
# mutate(soil_type = st) → adds soil type back as a column since cld() output doesn’t include it
# 
# select() → keeps only the columns needed for plotting: soil_type, treatment, and letters
# 
# Final mutate() → reorders factor levels so plots always show the treatments and soils in the correct order

```

```{r TIMEPOINT 1 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_INITIAL_mean, se_CH4_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL + 0.05*max(CH4_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - Create p-value annotations from model summaries}

CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 1 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_INITIAL_mean - se_CH4_ppm_INITIAL,
                    ymax = CH4_ppm_INITIAL_mean + se_CH4_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_I, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/pH/pH_CH4_conc_INITIAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 - linear models}

# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 2 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  emmeans_res <- CH4_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8") )  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 2 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_FINAL_mean, se_CH4_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL + 0.075*max(CH4_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - Create p-value annotations from model summaries}

CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_FINAL_mean - se_CH4_ppm_FINAL,
                    ymax = CH4_ppm_FINAL_mean + se_CH4_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_F, 
    aes(x = 1, 
        y = 7,
        # y = max(summ_exp$CH4_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,7.2)

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/pH/pH_CH4_conc_FINAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

# CO2 concentration analyses (without dry control)

```{r CO2 - summarizing data for concentration figures (initial & final) }

# Selecting only the experimental jars to include in the figure AND removing dry control jars 
exp_jars <- data %>%
  filter(
    jar_type == "Exp",
    pH_treatment != "Dry control"
  )

summ_exp <- exp_jars %>%  # Start with the 'exp_jars' dataset and pipe it forward
  group_by(soil_type, pH_treatment) %>% # Group the data by the factors 'soil_type' and 'pH_treatment'
  # All subsequent summary operations will be performed within these groups
 
   summarise(
    across(
      c(CO2_ppm_INITIAL, CO2_ppm_FINAL, CO2_ppm_I_timenorm, CO2_ppm_F_timenorm),  # Select the numeric columns to summarize: initial and final CO2 ppm
      list(
        mean = ~ mean(.x, na.rm = TRUE),  # Compute mean, ignoring NAs
        sd   = ~ sd(.x, na.rm = TRUE),    # Compute standard deviation, ignoring NAs
        n    = ~ n()                      # Count number of rows in the group
      ),
      .names = "{.col}_{.fn}"  
      # Automatically name the new summary columns using the pattern:
      # original column name + "_" + function name
      # e.g., CO2_ppm_INITIAL_mean, CO2_ppm_FINAL_sd
    ),
    .groups = "drop"  # Ungroup the data after summarising so further operations are not grouped
  ) %>% 
  
  # Calculate standard error and relevel factors
  mutate(
    se_CO2_ppm_INITIAL = CO2_ppm_INITIAL_sd / sqrt(CO2_ppm_INITIAL_n),  # Standard error = SD / sqrt(n) for initial CO2
    se_CO2_ppm_FINAL   = CO2_ppm_FINAL_sd / sqrt(CO2_ppm_FINAL_n),  # Standard error = SD / sqrt(n) for final CO2
    se_CO2_ppm_I_timenorm   = CO2_ppm_I_timenorm_sd / sqrt(CO2_ppm_I_timenorm_n),
    se_CO2_ppm_F_timenorm   = CO2_ppm_F_timenorm_sd / sqrt(CO2_ppm_F_timenorm_n),
    
    # Convert factors to ordered factors with specific levels
    soil_type    = factor(soil_type, levels = c("Unamended", "Ground Rock")),  
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  
  )

# write_csv(summ_exp,"summarized_data.csv")

```

## Timepoint 1 (initial)

```{r TIMEPOINT 1 - linear models}

# Fit models for each soil_type and store results
CO2_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CO2_ppm_INITIAL ~ pH_treatment, data = .x)   # fit linear model for CO2 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CO2_models

# Unamended soils model results 
summary(CO2_models_I$Unamended$model) # model summary 
anova(CO2_models_I$Unamended$model) # ANOVA table 
CO2_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CO2_models_I$`Ground Rock`$model) # model summary 
anova(CO2_models_I$`Ground Rock`$model) # ANOVA table 
CO2_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 1 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CO2_letters_I <- map_df(names(CO2_models_I), function(st) {
  emmeans_res <- CO2_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CO2_letters_I <- CO2_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CO2_ppm_INITIAL_mean, se_CO2_ppm_INITIAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CO2_ppm_INITIAL_mean + se_CO2_ppm_INITIAL + 0.05*max(CO2_ppm_INITIAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - Create p-value annotations from model summaries}

CO2_pvals_I <- map_df(names(CO2_models_I), function(st) {
  model <- CO2_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 1 - plot CH4 conc means and lm stats results}

# Plot
CO2_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CO2_ppm_INITIAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CO2_ppm_INITIAL_mean - se_CO2_ppm_INITIAL,
                    ymax = CO2_ppm_INITIAL_mean + se_CO2_ppm_INITIAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CO2_letters_I, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CO2_pvals_I, 
    aes(x = 3, 
        # y = 175500,
        y = max(summ_exp$CO2_ppm_INITIAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CO[2]~Concentrations~Timepoint~1),
       x = "pH Treatment",
       y = bquote(CO[2]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) 
# +
 # ylim(0,176000)

CO2_conc_means_I

ggsave(plot = CO2_conc_means_I, filename = "figures/pH/pH_CO2_conc_INITIAL_mean_yvary.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CO2_models,CO2_letters,CO2_pvals)

```


## Timepoint 2 (FINAL)

```{r TIMEPOINT 2 - linear models}

# Fit models for each soil_type and store results
CO2_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CO2_ppm_FINAL ~ pH_treatment, data = .x)   # fit linear model for CO2 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CO2_models

# Unamended soils model results 
summary(CO2_models_F$Unamended$model) # model summary 
anova(CO2_models_F$Unamended$model) # ANOVA table 
CO2_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CO2_models_F$`Ground Rock`$model) # model summary 
anova(CO2_models_F$`Ground Rock`$model) # ANOVA table 
CO2_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 2 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CO2_letters_F <- map_df(names(CO2_models_F), function(st) {
  emmeans_res <- CO2_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 2 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CO2_letters_F <- CO2_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CO2_ppm_FINAL_mean, se_CO2_ppm_FINAL),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CO2_ppm_FINAL_mean + se_CO2_ppm_FINAL + 0.05*max(CO2_ppm_FINAL_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 2 - Create p-value annotations from model summaries}

CO2_pvals_F <- map_df(names(CO2_models_F), function(st) {
  model <- CO2_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 2))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - plot CH4 conc means and lm stats results}

# Plot
CO2_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CO2_ppm_FINAL_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CO2_ppm_FINAL_mean - se_CO2_ppm_FINAL,
                    ymax = CO2_ppm_FINAL_mean + se_CO2_ppm_FINAL),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CO2_letters_F, 
    aes(x = pH_treatment, 
        y = letter_y, 
        label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CO2_pvals_F, 
    aes(x = 3, 
        y = 175500,
        # y = max(summ_exp$CO2_ppm_FINAL_mean)*1.1, 
        label = label),
    inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CO[2]~Concentrations~Timepoint~2),
       x = "pH Treatment",
       y = bquote(CO[2]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  ) +
  ylim(0,176000)

CO2_conc_means_F

ggsave(plot = CO2_conc_means_F, filename = "figures/pH/pH_CO2_conc_FINAL_mean.png", width = 8.5, height = 5)

# removing objects FINAL specific 

# rm(CO2_models,CO2_letters,CO2_pvals)

```


# OLD CODE 

## Combining timepoints 1 & 2 into one figrue 

```{r combined T1 & T2 CH4 conc figures}

# 1️⃣ Reshape summary data for plotting
summ_all <- summ_exp %>%
  dplyr::select(
    soil_type, pH_treatment,
    CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_mean,
    se_CH4_ppm_INITIAL, se_CH4_ppm_FINAL
  ) %>%
  pivot_longer(
    cols = -c(soil_type, pH_treatment),
    names_to = c("metric", "timepoint"),
    names_pattern = "(CH4_ppm|se_CH4_ppm)_(INITIAL|FINAL)"
  ) %>%
  mutate(
    type = if_else(str_detect(metric, "^se"), "se", "mean"),
    timepoint = recode(timepoint,
                       "INITIAL" = "Timepoint 1 (Initial)",
                       "FINAL"   = "Timepoint 2 (Final)")
  ) %>%
  dplyr::select(-metric) %>%
  pivot_wider(names_from = type, values_from = value)

# 2️⃣ Generate compact letters
generate_letters <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    emmeans_res <- CH4_models[[st]]$emmeans$emmeans
    cld_df <- multcomp::cld(emmeans_res, Letters = letters) %>%
      as.data.frame() %>%
      mutate(
        soil_type = st,
        timepoint = timepoint_label,
        letters = .group
      ) %>%
      dplyr::select(soil_type, pH_treatment, timepoint, letters)
  }) %>%
    mutate(
      soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),
      pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))
    )
}

CH4_letters_all <- bind_rows(
  generate_letters(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_letters(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "pH_treatment", "timepoint")) %>%
  mutate(letter_y = mean + se + 0.1 * max(mean))

# 3️⃣ Generate p-values
generate_pvals <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    model <- CH4_models[[st]]$model
    pval <- anova(model)$`Pr(>F)`[1]
    data.frame(
      soil_type = st,
      timepoint = timepoint_label,
      label = paste0("p = ", signif(pval, 3))
    )
  }) %>%
    mutate(soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")))
}

CH4_pvals_all <- bind_rows(
  generate_pvals(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_pvals(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "timepoint")) %>%
  group_by(soil_type, timepoint) %>%
  summarise(
    y = 7.7,
    # y = max(mean) + 0.15 * max(mean),
    label = unique(label),
    .groups = "drop"
  )

# 4️⃣ Plot combined figure
CH4_conc_combined_plot <- ggplot(summ_all, aes(x = pH_treatment, y = mean, fill = pH_treatment)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, size = 0.3) +
  geom_text(
    data = CH4_letters_all,
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_all,
    aes(x = Inf, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 0,
    size = 4.2,
    # fontface = "italic"
  ) +
  facet_grid(soil_type ~ timepoint) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62")) +
  labs(
    x = "pH Treatment",
    y = expression(paste(CH[4], " Concentration (ppm)")),
    title = expression(bold("Methane Concentrations Across Soil Types and Timepoints"))
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  ylim(0,8)

CH4_conc_combined_plot 

# Display and save
CH4_conc_combined_plot
ggsave(plot = CH4_conc_combined_plot,
       filename = "figures/10.13.25/CH4_conc_combined_timepoints.png",
       width = 10, height = 7)



```


# Time normalized CH4 concentration analyses (without dry control)

## Timepoint 1 (initial)

```{r TIMEPOINT 1 - linear models}

# Fit models for each soil_type and store results
CH4_models_I <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_I_timenorm ~ pH_treatment, data = .x)   # fit linear model for CH4 initial ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_I$Unamended$model) # model summary 
anova(CH4_models_I$Unamended$model) # ANOVA table 
CH4_models_I$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_I$`Ground Rock`$model) # model summary 
anova(CH4_models_I$`Ground Rock`$model) # ANOVA table 
CH4_models_I$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 1 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_I <- map_df(names(CH4_models_I), function(st) {
  emmeans_res <- CH4_models_I[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_I <- CH4_letters_I %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_I_timenorm_mean, se_CH4_ppm_I_timenorm),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_I_timenorm_mean + se_CH4_ppm_I_timenorm + 0.05*max(CH4_ppm_I_timenorm_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 1 - Create p-value annotations from model summaries}

CH4_pvals_I <- map_df(names(CH4_models_I), function(st) {
  model <- CH4_models_I[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 3))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 1 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_I <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_I_timenorm_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_I_timenorm_mean - se_CH4_ppm_I_timenorm,
                    ymax = CH4_ppm_I_timenorm_mean + se_CH4_ppm_I_timenorm),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_I, 
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(data = CH4_pvals_I, aes(x = 3, y = max(summ_exp$CH4_ppm_I_timenorm_mean)*1.1, label = label),
             inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Conc~Normalized~by~Incubation~Duration~(Days)~T1),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

CH4_conc_means_I

ggsave(plot = CH4_conc_means_I, filename = "figures/10.13.25/CH4_conc_normalized_INITIAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

## Timepoint 2 (final)

```{r TIMEPOINT 2 - linear models}

# Fit models for each soil_type and store results
CH4_models_F <- exp_jars %>% 
  group_by(soil_type) %>%                                # group data by soil type
  group_map(~ {                                           # apply function to each group
    lm_fit <- lm(CH4_ppm_F_timenorm ~ pH_treatment, data = .x)   # fit linear model for CH4 FINAL ppm vs pH_treatment
    list(                                                 # return both model and emmeans results
      model = lm_fit,                                     # store lm model
      emmeans = emmeans(lm_fit, pairwise ~ pH_treatment)  # get pairwise emmeans for pH_treatment
    )
  }, keep = TRUE) %>%                                     # keep soil_type info in output
  set_names(unique(exp_jars$soil_type))                   # name each list element by soil_type

 # all model output 
# CH4_models

# Unamended soils model results 
summary(CH4_models_F$Unamended$model) # model summary 
anova(CH4_models_F$Unamended$model) # ANOVA table 
CH4_models_F$Unamended$emmeans # returns both emmeans and pairwise objects              

# Unamended soils model results 
summary(CH4_models_F$`Ground Rock`$model) # model summary 
anova(CH4_models_F$`Ground Rock`$model) # ANOVA table 
CH4_models_F$`Ground Rock`$emmeans # returns both emmeans and pairwise objects              


```

```{r TIMEPOINT 2 - post-hoc pairwise comparisons and dynamically generate letters}

# Generate compact letter display for each soil type
CH4_letters_F <- map_df(names(CH4_models_F), function(st) {
  emmeans_res <- CH4_models_F[[st]]$emmeans$emmeans  # Extract emmeans table
  cld_df <- multcomp::cld(emmeans_res, Letters=letters)    # Generate compact letter display
  cld_df <- as.data.frame(cld_df)                   # Convert to standard data frame
  cld_df$soil_type <- st                            # Add soil type as a column
  cld_df$letters <- cld_df$.group                   # Rename .group column to 'letters'
  cld_df <- cld_df[, c("soil_type", "pH_treatment", "letters")]  # Keep only relevant columns
  return(cld_df)                                    # Return processed data frame
}) %>%
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),  # Relevel soil_type for plotting
    pH_treatment = factor(pH_treatment, levels = c("5", "7", "8") )  # Relevel pH_treatment
  )

```

```{r TIMEPOINT 2 - dynamically assign posthoc pairwise letter positions based on max value per soil type}

CH4_letters_F <- CH4_letters_F %>%
  left_join(summ_exp %>% dplyr::select(soil_type, pH_treatment, CH4_ppm_F_timenorm_mean, se_CH4_ppm_F_timenorm),
            by = c("soil_type", "pH_treatment")) %>%
  group_by(soil_type) %>%
  mutate(
    letter_y = CH4_ppm_F_timenorm_mean + se_CH4_ppm_F_timenorm + 0.05*max(CH4_ppm_F_timenorm_mean)
  ) %>%
  ungroup() %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - Create p-value annotations from model summaries}

CH4_pvals_F <- map_df(names(CH4_models_F), function(st) {
  model <- CH4_models_F[[st]]$model                # Extract the lm object
  pval  <- anova(model)$`Pr(>F)`[1]             # Extract the p-value for pH_treatment
  data.frame(
    soil_type = st,                              # Use the soil type name from the list
    label = paste0("p = ", signif(pval, 3))     # Format p-value
  )
}) %>% 
  mutate(
    soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock"))  # Relevel soil_type for plotting
  )

```

```{r TIMEPOINT 2 - plot CH4 conc means and lm stats results}

# Plot
CH4_conc_means_F <- ggplot(summ_exp, aes(x = pH_treatment, y = CH4_ppm_F_timenorm_mean, fill = pH_treatment)) +
  geom_col(position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = CH4_ppm_F_timenorm_mean - se_CH4_ppm_F_timenorm,
                    ymax = CH4_ppm_F_timenorm_mean + se_CH4_ppm_F_timenorm),
                width = 0.25, size = 0.3) +
  geom_text(
    data = CH4_letters_F, 
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(data = CH4_pvals_F, aes(x = 3, y = max(summ_exp$CH4_ppm_F_timenorm_mean)*1.1, label = label),
             inherit.aes = FALSE) +
  facet_wrap2(~soil_type, strip = strip_themed(
    background_x = elem_list_rect(fill = rep("gray95", 2)),
    text_x = elem_list_text(color = rep("black", 2))
  )) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62", "#8da0cb")) +
  theme_bw() +
  labs(title = bquote(CH[4]~Conc~Normalized~by~Incubation~Duration~(Days)~T2),
       x = "pH Treatment",
       y = bquote(CH[4]~(ppm)),
       fill = "Soil pH Treatment") +
  theme(
    plot.title = element_text(size = 20),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

CH4_conc_means_F

ggsave(plot = CH4_conc_means_F, filename = "figures/10.13.25/CH4_conc_normalized_FINAL_mean.png", width = 8.5, height = 5)

# removing objects initial specific 

# rm(CH4_models,CH4_letters,CH4_pvals)

```

## Combining timepoints 1 & 2 into one figrue 

```{r combined T1 & T2 CH4 conc figures}

# 1️⃣ Reshape summary data for plotting
summ_all <- summ_exp %>%
  dplyr::select(
    soil_type, pH_treatment,
    CH4_ppm_INITIAL_mean, CH4_ppm_FINAL_mean,
    se_CH4_ppm_INITIAL, se_CH4_ppm_FINAL
  ) %>%
  pivot_longer(
    cols = -c(soil_type, pH_treatment),
    names_to = c("metric", "timepoint"),
    names_pattern = "(CH4_ppm|se_CH4_ppm)_(INITIAL|FINAL)"
  ) %>%
  mutate(
    type = if_else(str_detect(metric, "^se"), "se", "mean"),
    timepoint = recode(timepoint,
                       "INITIAL" = "Timepoint 1 (Initial)",
                       "FINAL"   = "Timepoint 2 (Final)")
  ) %>%
  dplyr::select(-metric) %>%
  pivot_wider(names_from = type, values_from = value)

# 2️⃣ Generate compact letters
generate_letters <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    emmeans_res <- CH4_models[[st]]$emmeans$emmeans
    cld_df <- multcomp::cld(emmeans_res, Letters = letters) %>%
      as.data.frame() %>%
      mutate(
        soil_type = st,
        timepoint = timepoint_label,
        letters = .group
      ) %>%
      dplyr::select(soil_type, pH_treatment, timepoint, letters)
  }) %>%
    mutate(
      soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")),
      pH_treatment = factor(pH_treatment, levels = c("5", "7", "8"))
    )
}

CH4_letters_all <- bind_rows(
  generate_letters(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_letters(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "pH_treatment", "timepoint")) %>%
  mutate(letter_y = mean + se + 0.1 * max(mean))

# 3️⃣ Generate p-values
generate_pvals <- function(CH4_models, timepoint_label) {
  map_df(names(CH4_models), function(st) {
    model <- CH4_models[[st]]$model
    pval <- anova(model)$`Pr(>F)`[1]
    data.frame(
      soil_type = st,
      timepoint = timepoint_label,
      label = paste0("p = ", signif(pval, 3))
    )
  }) %>%
    mutate(soil_type = factor(soil_type, levels = c("Unamended", "Ground Rock")))
}

CH4_pvals_all <- bind_rows(
  generate_pvals(CH4_models_I, "Timepoint 1 (Initial)"),
  generate_pvals(CH4_models_F, "Timepoint 2 (Final)")
) %>%
  left_join(summ_all, by = c("soil_type", "timepoint")) %>%
  group_by(soil_type, timepoint) %>%
  summarise(
    y = 7.7,
    # y = max(mean) + 0.15 * max(mean),
    label = unique(label),
    .groups = "drop"
  )

# 4️⃣ Plot combined figure
CH4_conc_combined_plot <- ggplot(summ_all, aes(x = pH_treatment, y = mean, fill = pH_treatment)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, size = 0.3) +
  geom_text(
    data = CH4_letters_all,
    aes(x = pH_treatment, y = letter_y, label = letters),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  geom_label(
    data = CH4_pvals_all,
    aes(x = Inf, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 0,
    size = 4.2,
    # fontface = "italic"
  ) +
  facet_grid(soil_type ~ timepoint) +
  scale_fill_manual(values = c("darkgrey", "#66c2a5", "#fc8d62")) +
  labs(
    x = "pH Treatment",
    y = expression(paste(CH[4], " Concentration (ppm)")),
    title = expression(bold("Methane Concentrations Across Soil Types and Timepoints"))
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    strip.text = element_text(face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  ylim(0,8)

CH4_conc_combined_plot 

# Display and save
CH4_conc_combined_plot
ggsave(plot = CH4_conc_combined_plot,
       filename = "figures/10.13.25/CH4_conc_combined_timepoints.png",
       width = 10, height = 7)

```

